---
title: "Assignment 2 - Language Development in ASD - Part 1 - Explaining development"
author: "[Pernille]"
date: "[DATE]"
output: html_document
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
```

## NOTES:
```{r}
library(pacman)
p_load(tidyverse,lmerTest,MuMIn,sjPlot,lme4,reshape2, MuMIn)
```

Plot: Fitted values (predicted by model) against actual values 
broom.mixed
sjPlot or predict()


Get marginal (exp by fixed) + conditional (how much variance is exp with random effects also)
r.squaredGlMM() from MuMin

# TIPS
Use read_csv write_csv - it is 10 x faster than read.csv
Use tidyverse. Make code readable for both of us - no lon lines. 
# get overview
Readable - command space A - will space it out neatly

Readable things:
```{r}
#demo_train = demo_train %>%
#  rename(SUBJ = Child.ID,
#         VISIT = Visit)

#Consider making a function:
extractor <- function(df){
  
df <- df %>%
  mutate_at("VISIT", str_extract, "\\d") %>% # HVAD GØR DEN DER? 
  mutate(VISIT = as.integer(VISIT))
  
}

#Use the function: 
demo_train <- extractor(demo_train)
LU_train <- extractor(LU_train)
token_train <- extractor(token_train)

#Check that we have the same type: 
compare_df_cols(demo_train, LU_train, token_train) %>%
  tail()

# WHEN MERGING DATAFRAMES, DO ALL = TRUE
#Example solution 1 (although there should be something better): 
merged_data <- merge(demo_train, LU_train, all = T)
merged_data <- merge(merged_data, token_train, all = T) 

# SOLUTION AGAIN WITH FUNCTION
# example solution: 
selector2 <- function(df){
  
  df <- df %>%
    select(one_of(c("SUBJ", "VISIT", "Diagnosis", "Ethnicity",
           "Gender", "Age", "ADOS", "MullenRaw", "ExpressiveLangRaw",
           "Socialization", "MOT_MLU", "CHI_MLU", "types_MOT",
           "types_CHI", "tokens_MOT", "tokens_CHI"))) 

}

# run the function: 
demo_train <- selector2(demo_train)


# example solution for only having visit 1 so that we can try later and predict how it will be behave.
new_data <- merged_data %>%
  filter(VISIT == 1) %>%
  select(SUBJ,
         ADOS1 = ADOS, #new_name = old_name. 
         verbalIQ = ExpressiveLangRaw,
         nonVerbalIQ = MullenRaw,
         Socialization1 = Socialization)
  
data <- merge(new_data, merged_data, all = T)


```

```{r}
df <- alldata
#Preparing heatmap of correlations
sub_df <- df %>% select(-c(ID, ADOS, Socialization, nvIQ, vIQ))

sub_df <- na.omit(sub_df)

sub_df$Diagnosis <- as.factor(sub_df$Diagnosis)
sub_df$Diagnosis <- as.numeric(sub_df$Diagnosis)

sub_df$Ethnicity <- as.factor(sub_df$Ethnicity)
sub_df$Ethnicity <- as.numeric(sub_df$Ethnicity)

sub_df$Gender <- as.factor(sub_df$Gender)
sub_df$Gender <- as.numeric(sub_df$Gender)

# Functions
# Use correlation between variables as distance
reorder_cormat <- function(cormat) {
  dd <- as.dist((1 - cormat) / 2)
  hc <- hclust(dd)
  cormat <- cormat[hc$order, hc$order]
}

# Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat) {
  cormat[lower.tri(cormat)] <- NA
  return(cormat)
}

# Do the heatmap
cormat <- round(cor(sub_df),2)
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)

melted_cormat <- melt(upper_tri, na.rm = TRUE)

ggplot(melted_cormat, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1, 1),
    space = "Lab",
    name = "Pearson\nCorrelation"
  ) +
  theme_minimal() + # minimal theme
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    size = 10,
    hjust = 1
  )) +
  coord_fixed() + geom_text(aes(Var2, Var1, label = value),
                            color = "black",
                            size = 2) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    legend.justification = c(1, 0),
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal"
  ) +
  guides(fill = guide_colorbar(
    barwidth = 7,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

```

# Assignment 2

In this assignment you will have to discuss a few important questions (given the data you have). More details below. The assignment submitted to the teachers consists of:
- a report answering and discussing the questions (so we can assess your conceptual understanding and ability to explain and critically reflect)
- a link to a git repository with all the code (so we can assess your code)

Part 1 - Basic description of language development
- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced

- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
```{r}
alldata <- read_csv("cleanedData.csv")

# mean age by group for VISIT 1
alldata %>% 
  group_by(Diagnosis) %>% 
  filter(VISIT == 1) %>% 
  summarise(mean(Age, na.rm = T))

# Number of observations of each gender
alldata %>% 
  count(Gender)

            #1 = 310, 2 = 62

# Number of each diagnosis
alldata %>% 
  count(Diagnosis) # så vi har dataframet og så putter vi den ind i count og specifyer at vi gerne vil have det til at være for diagnosis

# ØVE - så vi vil gerne counte antal gender
alldata %>% 
  count(Diagnosis) #så behøver vi ikke lave dollartegnet, vi kan bare tage diagnosis - vi kan gøre det samme med de andre ting
alldata %>% 
  group_by(Diagnosis) %>% 
  count(Gender)

# vi kan lave et grouped df pr 1 visit
test <- alldata %>% 
  group_by(Diagnosis) %>% 
  filter(VISIT == 1)

# check names 
variable.names(alldata)

            #A = 176, B = 196

# n
nrow(alldata) #total 372 participants?  isits? 

# Mean socialization per diagnosis
alldata %>% 
  group_by(Diagnosis) %>% summarise( mean(Socialization))

# MLU as a function of Visit (diagnosis colored)
ggplot(alldata, aes(y = CHI_MLU, x = VISIT, color = Diagnosis),na.rm = TRUE)+
  geom_point()+
  geom_smooth(stat = "smooth", position = "identity", method = lm)
# Vi kan se at autistiske børn udvikler sig langsommere, de har en mere stump slope, hvor TD børn developer hurtigere


```

- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on? ARTICLE - reciprocal effects? 
```{r}
# Parental use of language as a function of VISIT
ggplot(alldata, aes(y = MOT_MLU, x = VISIT),na.rm = TRUE)+
  geom_point()+
  geom_smooth(stat = "smooth", position = "identity", method = lm)
```

- Include individual differences in your model of language development (in children). Identify the best model.
```{r}
# model with age*diagnosis
m <- lmer(CHI_MLU ~ 
            Age*Diagnosis + 
            (1+VISIT|ID), 
          alldata, REML = F)
summary(m)
r.squaredGLMM(m) #Marginal: 37,7%, Conditional: 82%

# model with visit*diagnosis
m1 <- lmer(CHI_MLU ~ 
            VISIT*Diagnosis + 
            (1+VISIT|ID), 
          alldata, REML = F)
summary(m1)
r.squaredGLMM(m1) #Marginal: 35,7%, Conditional: 81,5%

AIC(m) #560.82
AIC(m1) #572.46

```

Part 2 - Model comparison
- Discuss the differences in performance of your model in training and testing data
- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

Part 3 - Simulations to plan a new study
- Report and discuss a power analyses identifying how many new kids you would need to replicate the results

The following involves only Part 1.

## Learning objectives

- Summarize and report data and models
- Critically apply mixed effects (or multilevel) models
- Explore the issues involved in feature selection


# Quick recap
Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail:
i) relying on actual naturalistic language production,  ii) over extended periods of time.

We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

This RMarkdown file includes 
1) questions (see above). Questions have to be answered/discussed in a separate document that you have to directly submit on Blackboard.
2) A break down of the questions into a guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results in the doc for the teachers.

REMEMBER that you will have to have a github repository for the code and submit the answers to Blackboard without code (but a link to your github/gitlab repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

Before we get going, here is a reminder of the issues you will have to discuss in your report:

1- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
2- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
3- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
4- Include individual differences in your model of language development (in children). Identify the best model.

# Let's go

### Loading the relevant libraries

Load necessary libraries : what will you need?
- e.g. something to deal with the data
- e.g. mixed effects models
- e.g. something to plot with

```{r Load Libraries, include = FALSE}
library(pacman)
p_load(tidyverse,lmerTest,MuMIn,sjPlot,lme4,reshape2)

```

### Define your working directory and load the data
If you created a project for this class and opened this Rmd file from within that project, your working directory is your project directory.

If you opened this Rmd file outside of a project, you will need some code to find the data:
- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data, include = FALSE}
alldata <- read_csv("cleanedData.csv")
```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Socialization, Visit, Number of words used (token), Number of unique words used (type), mean length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups (ASD/TD) participants and whether the two groups are well matched.

```{r descriptive stats, include = FALSE}

# Making a grouped df
df_grp1 <- alldata %>% 
  group_by(Diagnosis) %>% 
  filter(VISIT == 1)

# Diagnosis as a factor, so I can see the levels
alldata$Diagnosis <- as.factor(alldata$Diagnosis)
levels(alldata$Diagnosis)

alldata %>% count(Diagnosis) #ASD = 176, TD = 196

# Mean age by group for VISIT 1
summarise(df_grp, "Mean Age Visit 1" = mean(Age, na.rm = T)) #ASD = 33, TD = 20.4

# Number of observations of each gender
df_grp %>% 
  count(Gender) #ASD F = 5, ASD M = 26, TD F = 6, TD M = 29

# Number of each diagnosis
df_grp %>% 
  count(Diagnosis) #ASD = 31, TD = 35

# n
nrow(df_grp) #total 66 participants

# Mean socialization per diagnosis (group by ID)
summarise(df_grp, "Mean Socialization Visit 1" = mean(Socialization, na.rm = T)) #ASD = 77.8, TD = 100

# -----------


# MLU as a function of Visit (diagnosis colored)
ggplot(alldata, aes(y = CHI_MLU, x = VISIT, color = Diagnosis),na.rm = TRUE)+
  geom_point()+
  geom_smooth(stat = "smooth", position = "identity", method = lm)
```

The sample included mostly young (<20) white males ...

[REPORT THE RESULTS]

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r ex2, include = FALSE}

```

How would you evaluate whether the model is a good model?

```{r ex2 evaluate, include = FALSE}

```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better.

```{r ex2 growth curve, include = FALSE}

```

Exciting right? Let's check whether the model is doing an alright job at fitting the data. Plot the actual CHI_MLU data against the predictions of the model fitted(model). 

```{r}

```

Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results
- A plot of your model's predictions (and some comments on whether the predictions are sensible)

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r ex3, include = FALSE}

```

Parent MLU is affected by ... but probably not ...
[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Victor and Byurakn.


```{r ex4, include = FALSE}

```

In addition to ..., the MLU of the children is also correlated with ...
Using AIC / nested F-tests as a criterium, we compared models of increasing complexity and found that ...

[REPORT THE RESULTS]